Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-06-14 analysis/paper/paper.Rmd
Writing conclusion and abstract

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/jas_reports.csl" # Insert path for the bib-style
---

@@

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @glørstad2011; @eigeland2015, 370]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, the relatively constrained geographical distribution of the sites hopefully counteracts some non-behavioural sources of variation.

@@

Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2013; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @kuhn2015; @riel-salvatore2008]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016; @Bicho2020; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181]. In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative [cf. @baxter1994, 100].

@@

As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data. A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)). While not as substantial, there is clear temporal patterning in the flint data as well. This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites. Slightly younger sites appear more impacted by core fragments and blades. The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades. Apart from the impact of core fragments, this must be considered very much in agreement with previous research [e.g. @solheim2017c, with references]. The comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g. @damlien2016; @eigeland2015; @solheim2020] . However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007]. The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf. @manninen2014].

@@

Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r =* `r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. The linear correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`). As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data. While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same. The relationship between mean site age and relative frequency of secondarily worked flint is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the x-axis, the volumetric density of flint is not temporally contingent (*r =* `r round(denscorflint, 2)`). As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.

@@

To get more directly at this temporal trend, a curation index based on VDL and RFSL was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @baxter2016; @crema2012; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot)A. Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This has stabilised by around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic. The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation associated mobility patterns. Figure \@ref(fig:simplot)B displays the result of running the same procedure on the flint data. The general pattern follows the same trajectory, but the result for some individual sites is markedly different. This is discussed below.

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE. This corresponds well with a chronological framework where the end of the Early Mesolithic, or Flake axe phase, is set to c. 8200 BCE [@reitan2016]. Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @fuglestvedt2012]. This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index. The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in Southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. Referring back to the increasing expediency in the curation data between c. 8200 and 7000 BCE, the Microlith phase could thus represent a transitional phase where migrating people and new living practices were propagating through societies in South-Eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.

The Microlith phase is followed the Pecked adze phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. As is evident from both the CA and the curation data, if we disregard Nedre Hobekk 2, the earliest of the assemblages treated here with this kind of compositional profile is site 40, Gunnarsrød 6b, dated to c. `r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[1]]*-1`--`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[2]]*-1` BCE [@carrasco2014]. The curation data remains stable from around 7000 BCE through the next typological transition at c. 5600 BCE, which, following @reitan2016, signifies the onset of the Nøstvet adze phase. While previously defined as having a slightly longer duration, the Nøstvet phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in south-eastern Norway [e.g. @jaksland2001; @lindblom1984], and has been explicitly linked to an expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit with the term being somewhat vaguely invoked [@eigeland2015, 127--130]. In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The curation index employed here clearly supports this notion, and suggests that the mobility patterns of the Nøstvet phase were well established in preceding periods.

The subsequent Transverse arrowhead phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged -and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021]. The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns in the material treated here. However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.

As it stands, however, the main hypotheses resulting from the analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and unform degrees of mobility, which then drop before leveling off at around 7000 BCE. These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA. Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@barton2011; @binford1980].

@@

The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway. This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories. These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages. Precisely what behavioural implication the development in the use of these types have, however, is less clear, but appears to follow a different, and more complex trajectory than that of curation, as operationalised here.

In my view this alignes well with the framework of @clark2017, suggesting that shifts in mobility patterns has caused this empirical pattern. However, the role of non-flint material, variability of flint quality. Furthermore, precisely how this mobility pattern has. Here Binfords model of briefly invoked, but there exists a wide range of possible configurations and mobility strategies that

>>>>

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/jas_reports.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as raw-material variability and excavation and classification practices, the sites chosen for analysis have all been excavated relatively recently, and have a constrained geographical distribution. Reaching the conclusion that these measures are applicable in a Norwegian Mesolithic setting. While this certainly holds potential for larger scale comparative studies, it also sets the stage for a more targeted and rigorous model evaluation outside an exploratory setting.

keywords: |
  Mesolithic Norway; Lithic assemblages; Multivariate statistics;
---

Keywords: `r rmarkdown::metadata$keywords`

@@

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @glørstad2011; @eigeland2015, 370]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

@@

Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2013; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016; @Bicho2020; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181]. In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative as compared to the CA [cf. @baxter1994, 100].

@@

As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data. A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)). While not as substantial, there is clear temporal patterning in the flint data as well. This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites. Slightly younger sites appear more impacted by core fragments and blades. The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades. Apart from the impact of core fragments, this must be considered very much in agreement with previous research [e.g. @solheim2017c, with references]. The comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g. @damlien2016; @eigeland2015; @solheim2020]. However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007]. The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf. @manninen2014].

@@

Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r =* `r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. The linear correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`). As variable non-flint availability and workability has also been suggested to potentially impact these dimensions [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data. While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same. The relationship between mean site age and relative frequency of secondarily worked flint is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the x-axis, the volumetric density of flint is not temporally contingent (*r =* `r round(denscorflint, 2)`). As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.

@@

To get more directly at this temporal trend, a curation index based on VDL and RFSL was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @baxter2016; @crema2012; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated summed posterior density estimates, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot)A. Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This has stabilised by around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic. The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in associated mobility patterns. Figure \@ref(fig:simplot)B displays the result of running the same procedure on the flint data. The general pattern follows the same trajectory, but the result for some individual sites is markedly different. This is discussed below.

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE. This corresponds well with a chronological framework where the end of the Early Mesolithic, or Flake axe phase, is set to c. 8200 BCE [@reitan2016]. Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012]. This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index. The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in Southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. Referring back to the increasing expediency in the curation data between c. 8200 and 7000 BCE, the Microlith phase could thus represent a transitional phase where migrating people and new living practices were propagating through societies in South-Eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.

The Microlith phase is followed by the Pecked adze phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. As is evident from both the CA and the curation data, if we disregard Nedre Hobekk 2, the earliest of the assemblages treated here with this kind of compositional profile is site 40, Gunnarsrød 6b, dated to c. `r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[1]]*-1`--`r sites[sites$site == "Gunnarsrød 6b", c("reported_start_bc", "reported_end_bc")][[2]]*-1` BCE [@carrasco2014]. The curation data remains stable from around 7000 BCE through the next typological transition at c. 5600 BCE, which, following @reitan2016, signifies the onset of the Nøstvet adze phase. While previously defined as having a slightly longer duration, the Nøstvet phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in south-eastern Norway [e.g. @jaksland2001; @lindblom1984], and has been explicitly linked to an expedient technological organisation [@glørstad2010, 161; @glørstad2011] --- albeit with the term being somewhat vaguely invoked [@eigeland2015, 127--130]. In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The curation index employed here clearly supports this notion, and suggests that the mobility patterns of the Nøstvet phase were well established in preceding periods.

The subsequent Transverse arrowhead phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021]. The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns in the material treated here. However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.

As it stands, the main hypotheses resulting from the present analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and uniform degrees of mobility, which then drop before levelling off at around 7000 BCE. These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA. Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@barton2011; @binford1980].

@@

The results of the CA align more or less with results of previous research in south-eastern Norway. This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories. These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages. Precisely what behavioural implication the development in the occurrences of the tool and debitage categories have are less clear, but appears to follow a different and more complex development over time than that of curation, as operationalised here.

The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns is the main driver behind this empirical pattern --- in line with the framework of @barton2011. Another perspective would be that this is not surprising at all [cf. @kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different. If its applicability and relevance for this setting is accepted, the relationship does hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond. Furthermore, the temporal trends associated with the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with Mesolithic south-eastern Norway. The explicit quantification does, however, offer the possibility of more formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.

Having reached the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting can only constitute a first analytical step. As @elster2015[12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached. This, however, can only constitute the absolute minimum of what is required of a proposed explanation. Subsequent steps should be to probe and challenge this explanatory framework.
